\phantomsection
\subsection*{Buddhist Causal Analysis: Craving Fulfilled, Suffering Renewed}
\addcontentsline{toc}{subsection}{Buddhist Causal Analysis}
\label{ssec:iii-buddhist-causal-analysis}

Early Buddhist teaching locates suffering not in moral disorder but in the causal structure of
craving itself. The \emph{Dhammapada} states the principle: ``From craving arises grief, from
craving arises fear; for one who is free of craving there is no grief or fear'' (Dhp. 216;
\parencite{BuddharakkhitaDhp1993}). The logic is causal, not punitive. Craving (\emph{taṇhā})
seeks to secure what is inherently unstable, and fulfillment does not resolve this---it
triggers the next cycle.

The \emph{Saṃyutta Nikāya}'s teaching of dependent origination
(\emph{paṭiccasamuppāda}) formalizes this: craving conditions clinging, clinging conditions
becoming, becoming conditions suffering (SN 12.2; \parencite[p.~536]{BodhiSN2000}). The key
insight is that craving, once gratified, does not terminate
but perpetuates the chain. Satisfaction is structurally impossible because the craving is not
for any particular thing but for permanence in what is impermanent. Fulfillment feeds the
chain rather than breaking it.

The Fire Sermon makes this visceral: ``The eye is burning \ldots\ burning with the fire of
lust, with the fire of hate, with the fire of delusion'' (SN 35.28;
\parencite[p.~1143]{BodhiSN2000}). Fire consumes its fuel. Craving, granted its object, does
not rest but continues to burn because the burning \emph{is} the craving. Fulfillment cannot
cool what is intrinsically aflame.

The remedy differs fundamentally from Biblical reordering. Rather than redirecting desire
toward a highest good, the Noble Eightfold Path aims at cessation (\emph{nirodha})---
extinguishing the fires through disenchantment, dispassion, and liberation
\parencite[pp.~45--50]{Rahula1959}. There is no reorientation toward an eternal Good but a
cooling, an unbinding (\emph{nibbāna}).

Coppola's \hyperref[scene:upriver-journey]{river journey} maps this phenomenology without
endorsing its metaphysics, and Willard's voiceover tracks the progression. Buddhist tradition
names the affect that arises from confronting impermanence \emph{saṃvega}---spiritual urgency,
the unsettling recognition that what one grasped cannot hold. Willard's early narration still
invests the mission with meaning: ``I wanted a mission, and for my sins they gave me one.
Brought it up to me like room service.'' The irony is defensive; the craving is intact. But
as the journey proceeds, each station exposes the structure.

The \hyperref[scene:playboy-show]{Playboy show} promises gratification and amplifies
agitation; the crew leaves more restless than before---craving fed, suffering renewed. After
the \hyperref[scene:sampan]{sampan massacre}, Willard's tone flattens: ``We cut them in half
with a machine gun and gave them a Band-Aid.'' \hyperref[scene:do-lung-bridge]{Do Lung Bridge}
literalizes samsaric repetition: rebuilt daily, destroyed nightly, progress without progress.
Willard observes the crew's hollow affect: ``They were mostly just out-of-it. This was going
on for twelve hours a day, every day.'' The \hyperref[scene:french-plantation]{French
plantation} dramatizes clinging (\emph{upādāna}): the colonists achieved their
object---land, dynasty, legacy---and will not let go even as fulfillment becomes living death.
They cling to what they won, and the clinging perpetuates suffering.

By the time Willard reaches \hyperref[scene:kurtz-compound]{Kurtz's compound}, his narration
has shed all anticipation. He no longer expects the mission to deliver meaning. This is
\emph{saṃvega} without the path that would follow it---disillusionment without liberation.
The mission itself, completed, does not bring rest but exposure. In Buddhist terms, each
``success'' is another link in the chain, and the chain's logic is revelation: what you
wanted cannot satisfy, because wanting is the problem.
